```{r include=FALSE, cache=FALSE}
set.seed(2466)

options(
  digits = 3,
  dplyr.print_max = 6,
  dplyr.print_min = 6
)

knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)

# Stamps plots with a tag 
# Idea from Claus Wilke's "Data Visualization" https://serialmentor.com/dataviz/
stamp <- function(
  tag = "Bad", tag_color = "#B33A3A", tag_size = 16, tag_padding = 1
)
{
  list(
    theme(
      plot.tag = element_text(color = tag_color, size = tag_size),
      plot.tag.position = "topright"
    ),
    labs(
      tag =
        str_pad(tag, width = str_length(tag) + tag_padding, side = "left")
    )
  )
}

```
# Time series

```{r, message=FALSE}
library(tidyverse)
library(lubridate)

# Data
gapminder <- read_rds("data/gapminder/data/combined_1950-2015_5yr.rds")
countries <- read_rds("data/gapminder/data/countries.rds")
life_expectancy <- 
  read_rds("data/gapminder/data/life_expectancy.rds") %>% 
  left_join(
    countries %>% select(iso_a3, region = region_gm4), by = "iso_a3"
  ) %>% 
  mutate(region = str_to_title(region))
wheat_prices <- read_rds("data/wheat-prices/wheat_prices.rds")
famines <- read_rds("data/famines/famines.rds")
```

## Mechanics

In the last chapter, we visualized the relationship between per capita GDP and life expectancy. You might have wondered how time fit into that association. In this chapter, we'll explore life expectancy and GDP over time. 

Just like with all other relationships, there are many different types of time series, and you'll want to highlight different aspects of the data for different time series. Sometimes, you'll be interested in a long-term trend, sometimes in short-term fluctuations, cyclical patterns, or even just individual values. In this chapter, we'll cover various strategies for dealing with these different scenarios. 

First, we'll talk about the mechanics of date scales, which can be useful for time series. 

### Date/time scales

Sometimes, your time series data will include detailed date or time information stored as a date, time, or datetime. For example, the `nycflights13::flights` variable `time_hour` is a datetime.

```{r}
nycflights13::flights %>% 
  select(time_hour)
```

When you map `time_hour` to an aesthetic, ggplot2 uses the scale function to datetimes: `scale_*_datetime()`. There is also `scale_*_date()` and `scale_*_time()` functions.

The date- and time-specific scale functions are useful because they create meaningful breaks and labels. 

```{r}
flight_counts_feb_week1 <- 
  nycflights13::flights %>% 
  filter(month == 2, day < 8) %>% 
  count(time_hour)

flight_counts_feb_week1 %>% 
  filter(day(time_hour) == 1) %>% 
  ggplot(aes(time_hour, n)) +
  geom_col()
```

Just like with the other scale functions, you can change the breaks using the `breaks` argument. `scale_*_date()` and `scale_*_datetime()` also include a `date_breaks` argument that allows you to supply the breaks in datetime units, like "1 month", "6 years", or "2 hours."

```{r}
flight_counts_feb_week1 %>% 
  filter(day(time_hour) == 1) %>% 
  ggplot(aes(time_hour, n)) +
  geom_col() +
  scale_x_datetime(date_breaks = "1 hour") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

Similarly, you _can_ change the labels using the `labels` argument, but `scale_*_date()` and `scale_*_datetime()` also include a `date_labels` function made for working with dates. `date_labels` takes the same formatting strings as functions like `ymd()` and `as_datetime()`. You can see a list of all formatting strings at `?strptime`.

Our plot above would be better if we just displayed the hours.

```{r}
flight_counts_feb_week1 %>% 
  filter(day(time_hour) == 1) %>% 
  ggplot(aes(time_hour, n)) +
  geom_col() +
  scale_x_datetime(date_breaks = "1 hours", date_labels = "%H") 
```


## Banking to 45

`gapminder` is the same Gapminder data you saw in the previous chapter. Previously, we investigated how life expectancy is associated with per capita GDP. Now, we'll ask a more obvious question: how has life expectancy changed over time?

```{r}
gapminder
```

First, let's just look at a single country: South Africa.

```{r}
south_africa <-
  gapminder %>% 
  filter(name == "South Africa")

south_africa %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() 
```

You can easily see how life expectancy has changed over time for South Africa. A line will visually connect the dots, making the trend easier to see and reinforcing the idea that there is an order to the dots.

```{r}
south_africa %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() +
  geom_line()
```

We'll explore this data further in the next section. First, we'll talk about a perceptual aspect of dealing with trends. 

Notice that, in the above plot, each line segment connecting two dots has its own angle. You judge this angle based on the lines orientation with respect to the x-axis. Changing the aspect ratio of the plot will change these angles.

Increasing the height to width ratio makes the angles steeper.

```{r}
south_africa %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() +
  geom_line() +
  coord_fixed(6)
```

Decreasing the height to width ratio makes the angles less steep.

```{r}
south_africa %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() +
  geom_line() +
  coord_fixed(.3)
```

Notice that life expectancy increased faster from 1980 to 1990 than it increased from 1950 to 1980. This feature is easier to see in the tall plot and harder to see in the wide plot. The reason for this difference is that, in the wide plot, the slopes of all line segments between 1950 and 1990 are more similar than the slopes of the same line segments in the tall plot. 

Cleveland, McGill, and McGill [@cleveland-1988] found that the aspect ratio that makes it easiest to discriminate different slopes is the one that makes the average slope of all line segments 45 degrees. This paper led to the "bank to 45" rule.

The default aspect ratio comes fairly close to adhering to this rule.

```{r}
south_africa %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() +
  geom_line() 
```

You don't need to calculate the angles of all your line segments and then mathematically work out the best aspect ratio. A second group of researchers also failed to replicate Cleveland, McGill, and McGill (1988), so it's not clear that the bank to 45 rule is really a "rule" (@talbot-2012).

However, it is useful to understand that aspect ratios can change the appearance of trends. Play around with your aspect ratio and notice which features of a trend are more visible, and choose an aspect ratio that highlights the features you think are most important.

## Trends

```{r}
south_africa %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() +
  geom_line() 
```

It looks like life expectancy started to fall around the time apartheid ended. We can add a reference line to check our hypothesis.

```{r}
south_africa %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1994, color = "blue")
```

We could remove the points and just use a line, but the points are helpful indicators of the actual data. Generally, the more points you have, the less important it is to use `geom_point()` in addition to `geom_line()`. For example, here's a plot of life expectancy for South Africa over time using the full `life_expectancy` data.

```{r}
south_africa_le <-
  life_expectancy %>% 
  filter(name == "South Africa")

south_africa_le %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1994, color = "blue")
```

The points are so close together that, for the most part, they form a line. Removing `geom_point()` doesn't affect the appearance of the plot very much.

```{r}
south_africa_le %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_line() +
  geom_vline(xintercept = 1994, color = "blue")
```

Including points is also important if your data is irregularly distributed, or if it is difficult to determine if peaks or values are outliers or real trends. 

The downward spike in 1918 is probably due to a Spanish flu outbreak. Why did life expectancy decline after the end of apartheid? One hypothesis is that data reporting procedures changed when the South African government changed. Maybe the apartheid government systematically undersampled non-white groups. Another hypothesis is that changes in government led to general upheaval, which somehow affected life expectancy. 

To further investigate, we could compare South Africa to its neighbors during this time period. When you compare multiple time series, it is especially important to use either lines and dots or just lines. 

```{r}
southern_africa_countries <-
  life_expectancy %>% 
  filter(name %in% c("South Africa", "Namibia", "Swaziland" ,"Lesotho")) %>% 
  filter(year > 1980, year < 2016)

 southern_africa_countries %>% 
  ggplot(aes(year, life_expectancy, color = name)) +
  geom_vline(xintercept = 1994, color = "blue") +
  geom_point() +
  scale_color_discrete(
    breaks = c("Namibia", "South Africa", "Swaziland", "Lesotho")
  ) +
  theme(legend.justification = "top")
```

Lines make it easier to visually connect different trends.

```{r}
 southern_africa_countries %>% 
  ggplot(aes(year, life_expectancy, color = name)) +
  geom_vline(xintercept = 1994, color = "blue") +
  geom_point() +
  geom_line() +
  scale_color_discrete(
    breaks = c("Namibia", "South Africa", "Swaziland", "Lesotho")
  ) +
  theme(legend.justification = "top")
```

In the previous chapter, you learned how to reorder and align the legend so that it is easier to match lines with legend labels. It's even easier to connect lines with labels if you directly label the lines on the plot. To do this, we'll use `ggrepel::geom_text_repel()`.

```{r}
southern_africa_countries %>% 
  ggplot(aes(year, life_expectancy, color = name)) +
  geom_vline(xintercept = 1994, color = "blue") +
  geom_point() +
  geom_line() +
  ggrepel::geom_text_repel(
    data = southern_africa_countries %>% filter(year == 2015),
    aes(label = name, x = year, y = life_expectancy), 
    size = 3, 
    nudge_y = .5,
    min.segment.length = 1,
    inherit.aes = FALSE
  ) +
  guides(color = "none") +
  theme(legend.justification = "top")
```

We also got rid of the legend by using `guides(color = "none")`.

Interestingly, all four countries experienced similar declines after 1994. This suggests that South Africa's decline was not related (at least entirely) to the end of apartheid. Next, we might wonder if this dip occurred in all of Africa. This brings us to the next section: visualizing distributions over time.

### Distributions over time

There are 54 countries listed under the "Africa" region in our data set. 

```{r}
africa_le <-
  life_expectancy %>% 
  filter(region == "Africa")

n_distinct(africa_le$name)
```

For each year, we'll have 54 different data points, which means we can't repeat our strategy of using `color` to encode different countries. 54 colors is far too many colors. 

One option is to use a smooth line to represent the general trend.

```{r}
africa_le %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_vline(xintercept = 1994) +
  geom_smooth()
```

(Note that `method = "loess"` is not a good idea here because we have `nrow(africa_le)` rows.)

You can see dip in the 1990s, but it starts earlier than South Africa's dip. 

Another option is to treat this problem as a multiple distribution problem. In that case, we can use the techniques covered in the _Distributions_ chapter. 

```{r}
africa_le %>% 
  filter(year > 1990, year < 2005) %>% 
  mutate(median_le = median(life_expectancy, na.rm = TRUE)) %>% 
  ggplot(aes(year, life_expectancy, group = year)) +
  geom_hline(aes(yintercept = median_le), color = "red") +
  geom_boxplot() +
  scale_x_continuous(breaks = seq(1990, 2005, 2))
```

Notice that we have to specify `group = year` because `year` is a continuous variable. `geom_boxplot()` doensn't assume it knows how to divide up continuous variables. 

The 1994 outlier is due to the Rwandan genocide.

```{r}
africa_le %>% 
  filter(year == 1994) %>% 
  top_n(-1, wt = life_expectancy)
```

If we exclude this point from view by zooming, the differences in medians become easier to see.

```{r}
africa_le %>% 
  filter(year > 1990, year < 2005) %>% 
  mutate(median_le = median(life_expectancy, na.rm = TRUE)) %>% 
  ggplot(aes(year, life_expectancy, group = year)) +
  geom_hline(aes(yintercept = median_le), color = "red") +
  geom_boxplot() +
  scale_x_continuous(breaks = seq(1990, 2005, 2)) +
  coord_cartesian(ylim = c(40, 80))
```

Again, you can see a slight dip in median life expectancies starting around 1995. 

This provides further evidence that the dip in South African life expectancies was not related to the end of apartheid. 

This dip is likely due the [HIV/AIDs pandemic](https://ourworldindata.org/grapher/share-of-the-population-infected-with-hiv?year=1995).

### Two response variables

So far, we've encoded time as position along the x-axis and a response variable, `life_expectancy`, along the y-axis. What if we want to want to encode a second response variable? It would be interesting to see how both life expectancy and per capita GDP have changed over time.

`geom_path()` is a useful way to represent a time series with two response variables. Unlike, `geom_line()`, which connects points in the order they appear along the x-axis, `geom_path()` connects points in the order they appear in the _data_. We can use this feature of `geom_path()` to represent a time series without actually plotting time along an axis.  

The following is a `geom_path()` plot of `gdp_per_capita` and `life_expectancy`. We've arranged the data by year so that the earliest years are plotted first.

```{r}
south_africa %>% 
  arrange(year) %>% 
  ggplot(aes(gdp_per_capita, life_expectancy)) +
  geom_path()
```

To read a `geom_path()` plot, you follow the line's path through the plot space. One problem with our current plot is that it doesn't indicate which end of the line is the starting point. We can add an arrow to our line indicating its direction.

```{r}
south_africa %>% 
  arrange(year) %>% 
  ggplot(aes(gdp_per_capita, life_expectancy)) +
  geom_path(arrow = arrow())
```

(See `?arrow` for arrow styling options.)

Now, it's clear that the line starts in the lower left corner. Both life expectancy and per capita GDP increased until around $1100. Then, life expectancy continued to grow, but per capita GDP shrank. Around $9025, this trend reversed. Life expectancy fell when per capita GDP increased. Finally, both started increasing again.

Another issue with our plot is that you can't match points on the path to specific years. One potential solution is to use color to encode `year`.

```{r}
south_africa %>% 
  arrange(year) %>% 
  ggplot(aes(gdp_per_capita, life_expectancy, color = year)) +
  geom_path() +
  scale_color_viridis_c()
```

The continuous color scale doesn't make it easy to accuratley map specific points to specific years, but you can get a rough sense of the timeline. Using `color` also means we don't need an arrow, since you can infer the direction of the path from the legend.

Another option is to label important points along the path.

```{r}
south_africa %>% 
  arrange(year) %>% 
  ggplot(aes(gdp_per_capita, life_expectancy)) +
  geom_path() +
  geom_text(
    data = south_africa %>% filter(year %in% c(1950, 1980, 1995, 2015)), 
    aes(label = year),
    nudge_x = -250
  )
```

### Multiple trend-types

All our life expectancy plots were relatively smooth. There were some outliers and downward trends, but, for the most part, life expectancy has been smoothly increasing. Often, however, you'll encounter data with multiple types of trends. 

The following plot shows the price of wheat over time.

```{r}
wheat_prices %>% 
  ggplot(aes(year, price)) +
  geom_point() +
  geom_line() 
```

There are two types of trends in this plot: a cyclical trend and a long-term downward trend.

When there are multiple types of trends, you'll need to decide which trend is most important. As you already learned, changing the aspect ratio affects your ability to discriminate different slopes. Changing the aspect ratio in this case will affect whether the cyclical trend or the long-term downward trend is more apparent.

```{r, fig.height = 6}
wheat_prices %>% 
  ggplot(aes(year, price)) +
  geom_point() +
  geom_line() +
  coord_fixed(.1)
```

A wide aspect ratio highlights the cyclical trends, but makes the long-term downward trend harder to spot.

```{r, fig.width = 10}
wheat_prices %>% 
  ggplot(aes(year, price)) +
  geom_point() +
  geom_line() +
  coord_fixed(1.2)
```

In this tall plot, the long-term downward trend is more apparent.

Let's make the plot slightly taller than the default to emphasize the downward trend starting in the early 1900s.

```{r}
wheat_prices %>% 
  ggplot(aes(year, price)) +
  geom_point() +
  geom_line() +
  coord_fixed(.6)
```

When your data has multiple types of trends, or a lot of noise, smoothing can be helpful.

```{r}
wheat_prices %>% 
  ggplot(aes(year, price)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "loess")
```

Notice that LOESS does not pick up on the cyclical trend, and only shows the long-term trend.

## Short-term fluctuations

In the mechanics section of this chapter, you saw the following plot.

```{r}
flight_counts_feb_week1 %>% 
  filter(day(time_hour) == 1) %>% 
  ggplot(aes(time_hour, n)) +
  geom_col()
```

You might wonder why we used `geom_col()` to represent a time series. Here's the same plot using `geom_line()` and `geom_point().

```{r}
flight_counts_feb_week1 %>% 
  filter(day(time_hour) == 1) %>% 
  ggplot(aes(time_hour, n)) +
  geom_point() +
  geom_line()
```

Both do a good job of showing that the most flights occurred in the morning, there's a dip in flight numbers around 11, and then another peak at 4pm. In this situation, though, it's helpful to see time as a discrete variable. We've counted the number of flights for each hour, and so it's useful to be able to connect a number of flights with a specific hour. Columns make it easier to connect a number of flights to an hour. 

Vertical segment plots using `geom_segment()` can also be helpful for some time series data. Say we want to understand what the first week in February looked like. `geom_point()` and `geom_line()` produce the following plot. 

```{r}
flight_counts_feb_week1 %>% 
  ggplot(aes(time_hour, y = n)) +
  geom_point() +
  geom_line() +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%a") 
```

You can see that each day is shaped similarly and that there are fewer flights on the weekend. However, you can't see that there are actually no flights for a couple hours each night, and it's difficult to connect points with specific hours.

```{r}
flight_counts_feb_week1 %>% 
  ggplot(aes(time_hour, xend = time_hour, y = 0, yend = n)) +
  geom_segment() +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%a") 
```

`geom_segment()` does a better job of showing the gaps between days. It's also easier to perceive each day as a group to compare against the others. Another advantage of `geom_segment()` is that we can use `color` to encode a categorical variable. 

```{r}
flight_counts_feb_week1 %>% 
  mutate(weekend = wday(time_hour, label = TRUE) %in% c("Sat", "Sun")) %>% 
  ggplot(aes(time_hour, xend = time_hour, y = 0, yend = n, color = weekend)) +
  geom_segment() +
  scale_x_datetime(date_breaks = "12 hours", date_labels = "%I %p") 
```

In this case, there's no long-term trend we're interested in. Instead, we want to understand short-term fluctuations, and care about individual values. In these situations, `geom_col()` and `geom_segment()` can be good options. 

## Individual values

Sometimes, you'll want to display time on the x-axis like a time series, but you won't actually care about displaying either short-term fluctuations or long-term trends. 

`famines` contains data on major famines across time.

```{r}
famines
```

Deaths due to famines have not changed much over time.

```{r}
famines %>%
  ggplot(aes(start, deaths_estimate)) +
  geom_point() +
  scale_y_log10()
```

The per capita GDP at which major famines occur has also not changed much over time.

```{r}
famines %>% 
  ggplot(aes(start, gdp_per_capita)) +
  geom_point() +
  scale_y_log10()
```

There's no trend, but we might be interested in the individual famines. Where did they occur? How many people died? 
  
We can use the strategy we talked about in _Continuous-continuous relationships_.

```{r, fig.width = 8}
famines %>% 
  arrange(desc(deaths_estimate)) %>% 
  ggplot(aes(start, gdp_per_capita, size = deaths_estimate, fill = region)) +
  geom_point(shape = 21, color = "black") +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  scale_size(range = c(1, 15)) +
  scale_y_log10()
```

The above plot only uses the `start` date, but we also have the length of the famines. We can treat the x-axis as representing year generally and encode the lenght of a line as the length of the famine.

```{r, fig.width = 8}
famines %>% 
  arrange(desc(deaths_estimate)) %>% 
  ggplot(
    aes(
      x = start, 
      xend = end, 
      y = deaths_estimate, 
      yend = deaths_estimate,
      color = region
    )
  ) +
  geom_segment(lineend = "round", size = 2) +
  scale_y_log10() +
  ggrepel::geom_text_repel(aes(label = name), size = 2.3, color  = "black") 
```


